---
title: "ST558, Project3"
author: "Jacob Press, Nataliya Peshekhodko"
date: "`r Sys.Date()`"
---



# Introduction

In this project we will read and analyze dataset [Diabetes health indicator dataset](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/) for each education level as specified by [EDUCA](https://www.icpsr.umich.edu/web/NAHDAP/studies/34085/datasets/0001/variables/EDUCA?archive=NAHDAP). **Level 1** (Never attended school or only kindergarten) and **Level 2** (Grades 1 - 8) will be combined. We will create different classification models for predicting the `Diabetes_binary` variable. The best model will be chosen based on `log loss` function.

Levels of educations based on EDUCA

```{r}
education_levels=list()
education_levels[['Never attended school or only kindergarten or Grades 1-8']] = '12'
education_levels[['Grades 9-11 (Some high school)']] = '3'
education_levels[['Grade 12 or GED (High school graduate)']] = '4'
education_levels[['College 1 year to 3 years (Some college or technical school)']] = '5'
education_levels[['College 4 years or more (College graduate)']] = '6'
```


This report is built for education level **`r names(unlist(education_levels))[unlist(education_levels) == params$education_level]`**.

# Packages

In order to achieve our goals, we will be using the following `R` packages.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(caret)
library(ggplot2)
library(corrplot)
```


  - `tidyverse` - is a collection of R packages, required for data transformation and manipulation
  - `caret` - required for training and evaluating machine learning models
  - `ggplot2` - required for for creating data visualizations and graphics
  - `corrplot` - required for correlation matrix visualizing


# Data

Reading data from `diabetes_binary_health_indicators_BRFSS2015.csv` file.

```{r, warning=FALSE, message=FALSE}
data = read_csv('./data/diabetes_binary_health_indicators_BRFSS2015.csv')
```


Checking for NA values
```{r}
sum(is.na(data))
```


There are no missing values in the data set.

Let's look at the head of data.
```{r}
head(data)
```


Combine Education levels `1` and `2` into one level `12`

```{r}
transformed <- data %>%
  mutate (Education = if_else(Education == 1 | Education == 2, 12, Education))
```

Sub-setting data for the selected education level:

```{r}
education_level = params$education_level

subset <- transformed %>%
  filter(Education == education_level)
```

Checking data structure:

```{r}
str(subset)
```


Variables in the data set:

  - **Diabetes_binary** - 0 = no diabetes, 1 = diabetes
  - **HighBP** - 0 = no high blood pressure, 1 = high blood pressure
  - **HighChol** - 0 = no high cholesterol, 1 = high cholesterol
  - **CholCheck** - 0 = no cholesterol check in 5 years,  1 = yes cholesterol check in 5 years
  - **BMI** - Body Mass Index
  - **Smoker** - Have you smoked at least 100 cigarettes in your entire life? 0 = no, 1 = yes
  - **Stroke** - (Ever told) you had a stroke. 0 = no,  1 = yes
  - **HeartDiseaseorAttack** - Coronary heart disease (CHD) or myocardial infarction (MI), 0 = no, 1 = yes
  - **PhysActivity** - Physical activity in past 30 days - not including job,  0 = no, 1 = yes
  - **Fruits** - Consume Fruit 1 or more times per day, 0 = no, 1 = yes
  - **Veggies** - Consume Vegetables 1 or more times per day, 0 = no 1 = yes
  - **HvyAlcoholConsump** - Heavy drinkers (adult men having more than 14 drinks per week and adult women having more than 7 drinks per week) 0 = no
  - **AnyHealthcare** - Have any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc. 0 = no 1 = yes
  - **NoDocbcCost** - Was there a time in the past 12 months when you needed to see a doctor but could not because of cost? 0 = no 1 = yes
  - **GenHlth** - Would you say that in general your health is: scale 1-5 1 = excellent 2 = very good 3 = good 4 = fair 5 = poor
  - **MentHlth** - Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how 
  - **PhysHlth** - Now thinking about your physical health, which includes physical illness and injury, for how many days during the past 30 
  - **DiffWalk** - Do you have serious difficulty walking or climbing stairs? 0 = no, 1 = yes
  - **Sex** - 0 = female,  1 = male
  - **Age** - 13-level age category, 1 = 18-24, 9 = 60-64,  13 = 80 or older
  - **Education** - Education level scale 1-6, 1 = Never attended school or only kindergarten,  2 = Grades 1 through 8
  - **Income** - Income scale scale 1-8,  1 = less than 10,000 dol,  5 = less than 35,000 dol,  8 = 75,000 dol or more
  

# Explanatory Data Analysis(EDA)

First, let's look at number of the records with Diabetes and without Diabetes for the selected education level:

```{r}
table (factor (subset$Diabetes_binary, labels = c("No diabet", "Diabet")) )
```

Let's look at `Age` distribution for the selected education level and check if all age groups are presented equally in the subset of data. 

```{r, message=FALSE}
ggplot(data = subset, aes(x = Age)) +
  geom_histogram(color = "black", fill = 'brown') +
  labs(title = "Histogram of Age groups distribution", 
       x = "Age group", 
       y = "Frequency")
```

Let's look at number of cases with Diabetes and without Diabetes for each age group for the selected education level.

```{r}
table(factor(subset$Diabetes_binary, labels = c("No diabet", "Diabet")), 
      factor(subset$Age, labels = c("Age 18 - 24", "Age 25 to 29", "Age 30 to 34", 
                                    "Age 35 to 39", "Age 40 to 44",
                                    "Age 45 to 49", "Age 50 to 54",
                                    "Age 55 to 59", "Age 60 to 64", 
                                    "Age 65 to 69", "Age 70 to 74",
                                    "Age 75 to 79", "Age 80 or older")) )
```

Let's check if number of cases with Diabetes and without Diabetes are equal for males and females for the selected subset of data.

```{r}
table(factor (subset$Diabetes_binary, labels = c("No diabet", "Diabet")), 
      factor(subset$Sex, labels = c("Female", "Male")))
```

Linear correlation between numeric variables allows to check which variables are correlated with target variable `Diabetes_binary` and could be used as predictors in the models.

```{r}
corrplot(cor(as.matrix(subset %>% select(-Education))), 
         type="upper", 
         tl.pos = "lt")
```

Number of cases with Diabetes and without Diabetes for each general health level.

```{r}
table(factor(subset$Diabetes_binary, labels = c("No diabet", "Diabet")), 
      factor(subset$GenHlth, labels = c("Excellent", "Very good", 
                                        "Good", "Fair", "Poor")) )
```
Number of cases with Diabetes and without Diabetes for high blood pressure and normal blood pressure patients.

```{r}
table(factor(subset$Diabetes_binary, labels = c("No diabet", "Diabet")), 
      factor(subset$HighBP, labels = c("No high BP", "High BP")) )
```

Number of cases with Diabetes and without Diabetes for high cholesterol and normal cholesterol patients.

```{r}
table(factor(subset$Diabetes_binary, labels = c("No diabet", "Diabet")), 
      factor (subset$HighChol, labels = c("No high chol", "High chol")))
```

BMI distribution for patients with Diabetes and without Diabetes for the selected education level.

```{r}
ggplot(subset, aes(x = as_factor(Diabetes_binary), 
                   y = BMI, 
                   fill = as_factor(Diabetes_binary))) +
  geom_boxplot() +
  labs(title = "BMI distribution for patients with and without diabetes", 
       x = "Diabetes", 
       y = "BMI") +
  scale_fill_manual(values = c("0" = "grey", "1" = "red"),
                    labels = c("0" = "Without Diabetes", "1" = "With Diabetes")) +
  labs(fill = "Diabetes Status")
```

Let's plot number of cases with Diabetes and Without Diabetes for each General Health level for the selected sunset of data.

```{r}
ggplot(subset, aes(x = as.factor(GenHlth), fill = as.factor(Diabetes_binary), group = Diabetes_binary)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Number of Cases with Diabetes and Without Diabetes by GenHlth Level",
    x = "GenHlth Level",
    y = "Number of Cases"
  ) +
  scale_fill_manual(
    values = c("0" = "grey", "1" = "red"),
    labels = c("0" = "Without Diabetes", "1" = "With Diabetes")
  ) + 
  labs(fill = "Diabetes Status")
```

Visualization of the proportion with and without diabetes grouped by Age for the given education level. It appears the proportion of with diabetes increases with age, which is not surprising.  

```{r}
ggplot(subset, aes(x = as.factor(Age), y = 1, fill = as.factor(Diabetes_binary), group = Diabetes_binary)) +
  geom_bar(position = "fill", stat = "identity") +
  labs(
    title = "Number of Cases with Diabetes and Without Diabetes by Age",
    x = "Age",
    y = "Number of Cases"
  ) +
  scale_fill_manual(
    values = c("0" = "grey", "1" = "red"),
    labels = c("0" = "Without Diabetes", "1" = "With Diabetes")
  ) +
  labs(fill = "Diabetes Status")
```

Visualization showing the distribution of `Diabetes_binary` by `BMI`.  

```{r}
ggplot(subset, aes(x = as.factor(Diabetes_binary), y = BMI, group = Diabetes_binary, fill = as.factor(Diabetes_binary))) +
 labs(
   title = "Violin Plot of Diabetes Status by BMI",
   x = "Diabetes Status",
   y = "BMI Scale"
 ) +
 geom_violin(trim = FALSE) +
 scale_fill_manual(
   values = c("0" = "grey", "1" = "red"),
   labels = c("0" = "Without Diabetes", "1" = "With Diabetes")
   ) +
 labs(fill = "Diabetes Status")
```

Visualization showing the counts of `Diabetes_binary` by `Income` for the given education level.  

```{r}
ggplot(subset, aes(x = as.factor(Income), fill = as.factor(Diabetes_binary), group = Diabetes_binary)) +
  geom_bar(position = "stack") +
  labs(
    title = "Number of Cases with Diabetes and Without Diabetes by Income",
    x = "Income",
    y = "Number of Cases"
  ) +
  scale_fill_manual(
    values = c("0" = "grey", "1" = "red"),
    labels = c("0" = "Without Diabetes", "1" = "With Diabetes")
  ) +
  labs(fill = "Diabetes Status")

```


# Modeling

Converting some of the variables to factors and checking dataset structure.

```{r}
names = c('HighBP' ,'HighChol', 
          'CholCheck', 'Smoker', 
          'Diabetes_binary', 'Stroke',
          'HeartDiseaseorAttack', 'PhysActivity',
          'Fruits', 'Veggies', 
          'HvyAlcoholConsump', 'Sex',
          'Age','Income', 'GenHlth', 
          'MentHlth', 'PhysHlth', 'DiffWalk',
          'AnyHealthcare', 'NoDocbcCost')
subset[,names] = lapply(subset[,names] , factor)
str(subset)
```


Splitting up data into training and validation datasets.

```{r}
set.seed(5)
trainIndex <- createDataPartition(subset$Diabetes_binary, p = .7, 
                                  list = FALSE, 
                                  times = 1)
train_data = subset[trainIndex, ]
val_data = subset[-trainIndex, ]
```


## Log loss

**Log loss**, also known as **logarithmic loss** or **cross-entropy loss**, is a common evaluation metric for binary classification models. It measures the performance of a model by quantifying the difference between predicted probabilities and actual values. Log-loss is indicative of how close the prediction probability is to the corresponding actual/true value, penalizing inaccurate predictions with higher values. **Lower log-loss** indicates **better** model performance.


  
Mathematical interpretation: Log Loss is the negative average of the log of corrected predicted probabilities for each instance.

$$log \ loss = -\frac{1}{N} \sum_{i=1}^N y_i log(p(y_i)) + (1-y_i)log(1-p(y_i))$$

$p(y_i)$ is the probability of $1$

  
$1-p(y_i)$ is the probability of 0


$y_i$ is the true binary outcome


We may prefer `log loss` to things like `accuracy` for several reasons:

  - **Probabilistic Evaluation** - Log loss considers probabilities, while accuracy only looks at final decisions
  - **Handles Imbalanced Data** - Log loss shows poor performance in imbalanced datasets
  - **Fair Model Comparison** - Log loss enables fair model comparisons and makes it easier to evaluate which model is performing better


## Logistic regression

**Logistic regression** is a statistical and machine learning model used for binary classification tasks. It's a type of regression analysis that's well-suited for predicting the probability of an observation belonging to one of two classes or categories.


  - Logistic regression is used when the response variable is binary, meaning it has two possible outcomes or classes. 
  - Logistic regression uses the `sigmoid` function to model the relationship between the features and the probability of the binary outcome. The logistic function has an S-shaped curve and maps any real-valued number to a value between 0 and 1. $p(x)=\frac{1}{1+e^{-(\beta_0+\beta_1x)}}$. ($p(x)$ is the probability of the dependent variable being 1)
  - The goal of logistic regression is to find the best-fitting model by estimating the coefficients $\beta_0$, $\beta_1$.
  This is typically done using a process called maximum likelihood estimation. The coefficients are adjusted to maximize the likelihood of the observed data given the model.
  
  
Creating lists to store model performances on train and validations data sets.

```{r}
models_performace_train = list()
models_performace_val = list()
logistic_regression_train = list()
```

### Fit Logistic regression model 1

Before we can fit logistic regression models, we need to transform response variable to the format `train` function expects using function `make.names`. 

```{r}
train_data$Diabetes_binary_transformed = train_data$Diabetes_binary
val_data$Diabetes_binary_transformed = val_data$Diabetes_binary

levels(train_data$Diabetes_binary_transformed) = make.names(levels(train_data$Diabetes_binary_transformed))
levels(val_data$Diabetes_binary_transformed) = make.names(levels(val_data$Diabetes_binary_transformed))
```

Fit logistic regression model with `HighChol`, `BMI` and `GenHlth` as predictors.

```{r, warning=FALSE}
train.control = trainControl(method = "cv", 
                              number = 5, 
                              summaryFunction=mnLogLoss,
                              classProbs = TRUE)

set.seed(83)
lr_model_1 = train(Diabetes_binary_transformed ~ 
                                   HighChol+
                                   BMI + 
                                   GenHlth, 
                                 data = train_data,
                                 method = "glm", 
                                 family="binomial",
                                 metric="logLoss",
                                 trControl = train.control
                                )
summary(lr_model_1)
```

Obtain `log loss` for train data set for logistic regression model #1.

```{r}
logistic_regression_train[['logistic_regression_model_1']] = lr_model_1$results$logLoss
print(paste("Obtained Log loss for for logistic regression model #1 on train dataset", 
            logistic_regression_train[['logistic_regression_model_1']]))
```


### Fit Logistic regression model 2

Fit logistic regression with second order `BMI`, `HighChol`, `HeartDiseaseorAttack` and interaction between `HighChol`and `HeartDiseaseorAttack` as predictors.

```{r, warning=FALSE}
train.control = trainControl(method = "cv", 
                              number = 5, 
                              summaryFunction=mnLogLoss,
                              classProbs = TRUE)

set.seed(8)
lr_model_2 = train(Diabetes_binary_transformed ~ 
                                   poly(BMI, 2) + 
                                   HighChol + HeartDiseaseorAttack+
                                   HighChol:HeartDiseaseorAttack,
                                 data = train_data,
                                 method = "glm", 
                                 family="binomial",
                                 metric="logLoss",
                                 trControl = train.control
                                )
summary(lr_model_2)
```


Obtain `log loss` for train data set for logistic regression model #2.

```{r}
logistic_regression_train[['logistic_regression_model_2']] = lr_model_2$results$logLoss
print(paste("Obtained Log loss for for logistic regression model #2 on train dataset", 
            logistic_regression_train[['logistic_regression_model_2']]))
```


### Fit Logistic regression model 3

Fit logistic regression with `Age`, `GenHlth`, `HighBP`, `HeartDiseaseorAttack` and second order `BMI` as predictors.

```{r, warning = FALSE}
train.control = trainControl(method = "cv", 
                              number = 5, 
                              summaryFunction=mnLogLoss,
                              classProbs = TRUE)

set.seed(10)
lr_model_3 = train(Diabetes_binary_transformed ~ Income+
                     Age+GenHlth+
                     HighBP+
                     HeartDiseaseorAttack+
                     poly(BMI, 2),
                   data = train_data,
                   method = "glm", 
                   family="binomial",
                   metric="logLoss",
                   trControl = train.control
                   )
summary(lr_model_3)
```

Obtain `log loss` for train data set for logistic regression model #3.

```{r}
logistic_regression_train[['logistic_regression_model_3']] = lr_model_3$results$logLoss
print(paste("Obtained Log loss for for logistic regression model #3 on train dataset", 
            logistic_regression_train[['logistic_regression_model_3']]))
```


### The best performed logistic regression model

Now we can choose the best performed model based on train dataset performance among logistic regression models.

```{r}
best_lr_model = names(logistic_regression_train)[which.min(unlist(logistic_regression_train))]
```

The best performed logistic regression model is model **`r best_lr_model`**.

Save the logistic regression model performance on training dataset.

```{r}
if (best_lr_model == "logistic_regression_model_1") {
  lr_model = lr_model_1
} else if (best_lr_model == "logistic_regression_model_2") {
  lr_model = lr_model_2
} else {
  lr_model = lr_model_3
}

models_performace_train[['logistic_regression_model']] = lr_model$results$logLoss

print(paste("Log Loss for logistic regression model for training dataset:", 
            models_performace_train[['logistic_regression_model']]))
```


In order to obtain `log loss` metric for validation dataset, let's create custom function for log loss calculation.

```{r}
calculateLogLoss <- function(predicted_probabilities, true_labels) {
  predicted_probabilities = pmax(pmin(predicted_probabilities, 1 - 1e-15), 
                                 1e-15)

  log_loss <- -mean(true_labels * log(predicted_probabilities) + 
                      (1 - true_labels) * log(1 - predicted_probabilities))
  return(log_loss)
}
```

Calculate and save logistic regression model performance on validation dataset.

```{r}
val_predictions = predict(lr_model, 
                             newdata = val_data %>% select(-Diabetes_binary), 
                             type = "prob")

predicted_prob_class1 = val_predictions[, 1]
true_labels = as.integer(as.character(val_data$Diabetes_binary))

log_loss_val_lr_model = calculateLogLoss(predicted_prob_class1, true_labels)
print(paste("Log Loss for logistic regression model for validation dataset:", 
            log_loss_val_lr_model))

models_performace_val[["logistic_regression_model"]] = log_loss_val_lr_model
```


## LASSO logistic regression

`LASSO (Least Absolute Shrinkage and Selection Operator) logistic regression` is a statistical method that combines logistic regression with LASSO regularization. It is used for binary classification problems where you want to predict the probability of an event occurring based on various predictor variables.

$$\sum_{i=1}^n(y_i-\sum_{j}x_{ij}\beta_j)^2 + \lambda \sum_{j=1}^p |\beta_j|$$


  -  `LASSO logistic regression` models the probability of an event using the logistic function. It models the log-odds of the event as a linear combination of predictor variables. The logistic function is used to transform the linear combination into probabilities.
  - `LASSO` adds a regularization term to the logistic regression model. The regularization term is a penalty based on the absolute values of the model coefficients (L1 regularization). This penalty encourages some of the coefficient values to become **exactly zero**, effectively performing feature selection.
  - `LASSO` regularization promotes sparsity in the model. It can automatically select a subset of the most relevant predictor variables by setting the coefficients of irrelevant variables to zero. This helps to reduce overfitting and build more interpretable models.
  - The degree of regularization is controlled by a hyper parameter denoted as $\lambda$.
  

Using Lasso models in logistic regression offers benefits such as automatic feature selection, better generalization to new data, model stability, improved interpretability, handling multicollinearity, and variable importance assessment.


### Fit and validate LASSO logistic regression


```{r, warning=FALSE}
train.control = trainControl(method = "cv",
                              number = 5, 
                              summaryFunction=mnLogLoss,
                              classProbs = TRUE)


set.seed(2)

# Limiting number of features due to performance issues
lasso_log_reg<-train(#Diabetes_binary ~., 
                   Diabetes_binary ~ HighBP + HighChol + BMI + Smoker + AnyHealthcare + GenHlth + Age + Sex,
                   data = select(train_data, -Diabetes_binary_transformed),
                   method = 'glmnet',
                   metric="logLoss",
                   tuneGrid = expand.grid(alpha = 1, 
                                          lambda=seq(0, 1, by = 0.25))
)
lasso_log_reg$results
```

Obtained the best tuning parameter $\lambda$ value is

```{r}
lasso_log_reg$bestTune$lambda
```

Plot obtained accuracy for different $\lambda$ values.

```{r}
plot(lasso_log_reg)
```


Calculate log loss for train data set


```{r}
train_predictions = predict(lasso_log_reg, 
                             newdata = train_data %>% select(-Diabetes_binary), 
                             type = "prob")

predicted_prob_class1 = train_predictions[, 1]
true_labels = as.integer(as.character(train_data$Diabetes_binary))

log_loss_train_lasso = calculateLogLoss(predicted_prob_class1, true_labels)
print(paste("Log Loss for lasso regression model for train dataset", 
            log_loss_train_lasso))

models_performace_train[["lasso"]] = log_loss_train_lasso
```

Calculate log loss for validation data set

```{r}
val_predictions = predict(lasso_log_reg, 
                             newdata = val_data %>% select(-Diabetes_binary), 
                             type = "prob")

predicted_prob_class1 = val_predictions[, 1]
true_labels = as.integer(as.character(val_data$Diabetes_binary))

log_loss_val_lasso = calculateLogLoss(predicted_prob_class1, true_labels)
print(paste("Log Loss for lasso regression model for validation dataset", 
            log_loss_val_lasso))

models_performace_val[["lasso"]] = log_loss_val_lasso
```


## Classification tree model

A Classification tree model is a supervised machine learning model used to predict group membership. It has a hierarchical tree structure consisting a root node, branches, internal nodes, and leaf nodes. Classification trees are used when the target variable is categorical. One benefit to classification trees is they are intuitive and usually easy to explain.  

Here is a break down of the tree structure:

  - **Root Node** - The beginning node on the graph.
  - **Branches** - The arrows connecting the nodes.
  - **Internal Nodes** - A non-leaf node denoting a test on an attribute.
  - **Leaf Nodes** - The terminal node displaying the classification.

### Fit and validate classification tree model

```{r}
train_control <- trainControl(method = "cv",
                              summaryFunction=mnLogLoss,
                              classProbs = TRUE,
                              number = 5)
set.seed(1122)
tree_model <- train(Diabetes_binary_transformed ~ ., 
                data = select(train_data, -Diabetes_binary), 
                method = "rpart",
                trControl = train_control,
                metric="logLoss",
                tuneGrid = data.frame(cp=seq(0,.022, by = .001))
                )

tree_model$results
```

```{r}
plot(tree_model)
```

Obtain log loss for train data set

```{r}
models_performace_train[["classification_tree"]] = min(tree_model$results$logLoss)
print(paste("Log Loss for classification tree model for training dataset:", 
            models_performace_train[["classification_tree"]]))
```

Calculate log loss for validation data set

```{r}
val_predictions = predict(tree_model, 
                             newdata = val_data %>% select(-Diabetes_binary_transformed), 
                             type = "prob")

predicted_prob_class1 = val_predictions[, 1]
true_labels = as.integer(as.character(val_data$Diabetes_binary))

log_loss_val_tree = calculateLogLoss(predicted_prob_class1, true_labels)
print(paste("Log Loss for classification tree for validation dataset", 
            log_loss_val_tree))

models_performace_val[["classification_tree"]] = log_loss_val_tree
```

## Random forest model

A Random Forest classification model is a supervised machine learning model used for classification tasks. It is an ensemble of multiple decision trees, where each tree predicts the class label of an input based on a set of features. The final prediction in a Random Forest is determined through a combination of predictions from individual decision trees, often using **majority voting** for classification tasks.


Random Forest might be chosen over a basic Classification Tree for several reasons:

  - **Generalization** - Random Forest typically offers better generalization to new, unseen data. It reduces the risk of overfitting, which is a common issue with basic Classification Trees.
  - **Higher Accuracy** - Random Forest often provides higher accuracy because it combines multiple decision trees. The majority voting from these trees leads to a more reliable and accurate classification.
  - **Robustness to Noise** - Basic Classification Trees are sensitive to noise in the data, which can lead to overfitting. Random Forest, through ensemble learning, is more robust to noise and outliers.
  - **Reduced Variance** - A basic Classification Tree can vary significantly with small changes in the training data. Random Forest reduces this variance because the ensemble of trees accounts for different sources of variance.
  - **Feature Selection** - Random Forest provides a measure of feature importance. It can help identify which features are most relevant for making predictions. This feature selection is especially valuable when dealing with high-dimensional data.
  

There are several hyperparameters which could be fine-tuned during random forest model training. Some which have the biggest affect to model performance:

  - **mtry** - Number of predictor variable randomly selected to be sampled at each split of the tree.  It controls the level of feature randomness in each tree. A smaller mtry may reduce overfitting, while a larger mtry can lead to better diversity among trees.
  - **ntree** - The number of decision trees (or "trees") to be grown in the forest. Increasing the number of trees can improve the model's accuracy, but it also increases computational cost.

  
### Fit and validate random forest model
  
```{r}
train_control <- trainControl(
  method = "cv",   
  number = 5,
  summaryFunction=mnLogLoss,
  classProbs = TRUE
)

# Limiting number of the features due 
# to performance issues with random forest algorithm.
# Ideally we should try to use all features:  Diabetes_binary_transformed ~ ., 
set.seed(11)
rf_model = train(
  Diabetes_binary_transformed ~ HighChol+
                                BMI + 
                                GenHlth+
                                HeartDiseaseorAttack+
                                Age+
                                Income,
  data = select(train_data, -Diabetes_binary),
  method = "rf",
  metric="logLoss",
  tuneGrid = data.frame(mtry = c(1:4)), 
  trControl = train_control
)

rf_model$results
```

```{r}
plot(rf_model)
```



Obtain log loss for train data set


```{r}
models_performace_train[["random_forest"]] = min(rf_model$results$logLoss)
print(paste("Log Loss for random forest model for training dataset:", 
            models_performace_train[["random_forest"]]))
```

Calculate log loss for validation data set

```{r}
val_predictions = predict(rf_model, 
                             newdata = val_data %>% select(-Diabetes_binary), 
                             type = "prob")

predicted_prob_class1 = val_predictions[, 1]
true_labels = as.integer(as.character(val_data$Diabetes_binary))

log_loss_val_rf = calculateLogLoss(predicted_prob_class1, true_labels)
print(paste("Log Loss for random forest model for validation dataset", 
            log_loss_val_rf))

models_performace_val[["random_forest"]] = log_loss_val_rf
```


## New model - Support Vector Machine

Support Vector Machine(SVM) is a supervised machine learning algorithm that is used for both classification and regression tasks. It is a powerful and versatile algorithm known for its ability to handle complex decision boundaries and high-dimensional data. SVM works by finding the *optimal hyperplane* that best separates data points into different classes or predicts a continuous target variable.

Main components of SVM:

  - **Hyperplane** - SVM's core concept is to find the optimal hyperplane that maximizes the margin between two classes in a data set. The hyperplane is the decision boundary that separates data points into different classes. In two dimensions, it's a line; in higher dimensions, it's a hyperplane.
  - **Support Vectors** - Support Vectors are the data points that are closest to the decision boundary, or hyperplane. These support vectors play a crucial role in determining the position and orientation of the hyperplane.
  - **Margin** - The margin is the distance between the decision boundary (hyperplane) and the closest support vectors. SVM aims to maximize this margin, as it represents the separation between classes. The larger the margin, the better the model's generalization.
  - **Kernel Trick** - SVM can handle both linearly separable and non-linearly separable data. The kernel trick allows SVM to transform data into higher-dimensional space, making it possible to find linear separation in this transformed space. Common kernel functions include linear, polynomial, radial basis function (RBF), and sigmoid.
  - **C Parameter** - SVM has a hyper parameter called C, which controls the trade-off between maximizing the margin and minimizing the classification error. Smaller C values lead to a larger margin but may allow some mis-classification, while larger C values lead to a smaller margin with fewer mis-classifications.
  
  
### Fit and validate support vector machine model  
  
```{r, warning=FALSE}
train_control = trainControl(
  method = "cv",
  number = 5,
  classProbs =  TRUE,
  summaryFunction=mnLogLoss
)

# limiting number of hyperparameters due to the
# performance issues
svm_grid = expand.grid(
  sigma = c(0.1),
  C = c(1, 10)
)

# Limiting number of features due to the performance 
# issues. Ideally, we should try to use all predictors: Diabetes_binary_transformed ~ .
svm_model = train(
  Diabetes_binary_transformed ~ HighChol+
                                BMI + 
                                GenHlth,
  data = select(train_data, -Diabetes_binary),
  method = "svmRadial",
  trControl = train_control,
  metric="logLoss",
  tuneGrid = svm_grid
)
svm_model$results
```


```{r}
plot(svm_model)
```


```{r}
svm_model$bestTune
```

Obtain log loss for train dataset.


```{r}
models_performace_train[["svm"]] = min(svm_model$results$logLoss)
print(paste("Log Loss for support vector machine model for training dataset", 
            models_performace_train[["svm"]]))
```


Calculate log loss for validation dataset.

```{r}
val_predictions = predict(svm_model, 
                             newdata = val_data %>% select(-Diabetes_binary_transformed), 
                             type = "prob")

predicted_prob_class1 = val_predictions[, 1]
true_labels = as.integer(as.character(val_data$Diabetes_binary))

log_loss_val_svm = calculateLogLoss(predicted_prob_class1, true_labels)
print(paste("Log Loss for support vector machine model for validation dataset", 
            log_loss_val_svm))

models_performace_val[["svm"]] = log_loss_val_svm
```

## New model - Naive Bayes

The naive Bayes model is a popular supervised machine learning algorithm used for classification. An advantage of a naive bayes model is it can require a smaller portion of training data to predict classification. But naive bayes may be out performed by other models such as boosted trees and random forests. A break down of Bayes theorem below:

$$ P(c|x) = \frac{P(x|c)*P(c)}{P(x)}$$
Where:

  - $P(c|x)$ is the posterior probability of C (class membership) given x (predictor).
  - $P(c)$ prior probability of class membership.
  - $P(x|c)$ is the probability of the predictor given class membership.
  - $P(x)$ is the prior probability of the predictor.


### Fit and validate Naive Bayes model

```{r}
train_control = trainControl(
  method = "cv",
  number = 5,
  summaryFunction=mnLogLoss,
  classProbs =  TRUE
)

nb_grid <- expand.grid(
  usekernel = TRUE,
  fL = 1,
  adjust = seq(0.5, 1.5, by = 0.5)
)

nb_model = train(
  Diabetes_binary_transformed ~ ., 
  data = select(train_data, -MentHlth, -PhysHlth, -Diabetes_binary),
  method = "nb",
  trControl = train_control,
  tuneGrid = nb_grid,
  metric="logLoss"
)

nb_model$results
```

```{r}
plot(nb_model)
```


Obtain log loss for train dataset.


```{r}
models_performace_train[["Naive Bayes"]] = min(nb_model$results$logLoss)
print(paste("Log Loss for naive bayes model for training dataset", 
            models_performace_train[["Naive Bayes"]]))
```


Calculate log loss for validation dataset.

```{r}
val_predictions = predict(nb_model, 
                             newdata = val_data %>% select(-Diabetes_binary_transformed), 
                             type = "prob")

predicted_prob_class1 = val_predictions[, 1]
true_labels = as.integer(as.character(val_data$Diabetes_binary))

log_loss_val_nb = calculateLogLoss(predicted_prob_class1, true_labels)
print(paste("Log Loss for naive bayes model for validation dataset", 
            log_loss_val_nb))

models_performace_val[["Naive Bayes"]] = log_loss_val_nb
```

All models performance on training dataset based on `log loss` metric:

```{r}
models_performace_train
```

All models performance on validation dataset based on `log loss` metric:
```{r}
models_performace_val
```

The best performed model based on train data set is  **`r names(models_performace_train)[which.min(unlist(models_performace_train))]`**.


The best performed model based on validation data set is **`r names(models_performace_val)[which.min(unlist(models_performace_val))]`**.



# Summary

In this report we analyze subset of dataset [Diabetes health indicator dataset](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/) for educational level - **`r names(unlist(education_levels))[unlist(education_levels) == params$education_level]`**. We fit and validate six different machine learning models. Based on performance on validation dataset the best model is **`r names(models_performace_val)[which.min(unlist(models_performace_val))]`**.
